---
title: "Mortality in South Africa"
output: 
  flexdashboard::flex_dashboard:
    theme: lumen
    orientation: columns
    vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(southafricastats)
library(dplyr)
library(tidyr)
library(leaflet)
library(ggplot2)
library(DT)

totals <- population_zaf %>% 
    filter(year == 2013) %>% 
    select(province, total)

compare_provinces <- mortality_zaf %>%
    left_join(totals) %>%
    mutate(rate = deaths / total * 1e3) %>%
    group_by(province, indicator) %>%
    summarise(rate = mean(rate)) %>%
    ungroup %>%
    left_join(totals) %>%
    spread(indicator, rate)

mortality_zaf <- mortality_zaf %>%
    filter(indicator != "All causes")

```

Column {.sidebar}
==================================================

This app explores mortality data from the 
[South Africa Data Portal](http://southafrica.opendataforafrica.org/).

Use the input below to select a cause of death to explore.

```{r}
selectInput(inputId = "indicator", 
            label = "Choose a cause of death:",
            choices = sort(unique(mortality_zaf$indicator)),
            selected = "Cerebrovascular diseases (I60-I69)")
```

Map {data-icon="fa-map-marker"}
==================================================

### Where is the mortality rate highest?

```{r}
selected_df <- reactive({
    subset_df <- compare_provinces[,c(1, which(colnames(compare_provinces) == input$indicator))]
    colnames(subset_df) <- c("province", "indicator")
    population_zaf %>%
      filter(year == 2013) %>%
      left_join(subset_df) %>%
      mutate(indicator = indicator / sum(indicator, na.rm = T))
})

renderLeaflet({
  leaflet(selected_df()) %>%
    addProviderTiles("CartoDB.Positron") %>%
    addCircles(lng = ~longitude,
               lat = ~latitude,
               radius = ~sqrt(indicator) * 3e5,
               weight = 2.5,
               color = "magenta",
               popup = ~province)
})
```

Comparing provinces {data-icon="fa-list"}
==================================================

### How does the selected mortality rate compare to the overall mortality rate?

```{r}

scatterplot_df <- reactive({
    subset_df <- compare_provinces[, c(1:2, 
                                       which(colnames(compare_provinces) == "All causes"),
                                       which(colnames(compare_provinces) == input$indicator))]
    colnames(subset_df) <- c("province", "total", "selected_x", "selected_y")
    subset_df
})

renderPlot({
  ggplot(scatterplot_df(),
       aes(`All causes`, indicator,
           label = province)) +
  geom_point(aes(size = total),
             alpha = 0.8, color = "midnightblue") +
  geom_text(vjust = 2.5) +
  theme_minimal() + 
  labs(x = "All causes",
       y = input$indicator)
})
```

Changes in time {data-icon="fa-area-chart"}
==================================================

### How have the number of deaths changed over time?

```{r}
renderPlot({
    mortality_zaf %>%
        filter(indicator == input$indicator) %>%
        ggplot(aes(year, deaths, color = province)) +
        geom_line(alpha = 0.8, size = 1.5) +
        theme_minimal(base_size = 18) +
        labs(x = NULL, y = "# of deaths per year")
})
```

Changes in time {data-icon="fa-table"}
==================================================

### Check out the data here

```{r}
renderDataTable({
      mortality_zaf %>%
      filter(indicator == input$indicator)
}
)

```


